A Deeper Analysis into the Trends and Factors of the US's Declining Birth Rates

### 2024-02-13 paper/paper.qmd
added figure 5

---
title: "A Deeper Analysis into the Trends and Factors of the US's Declining Birth Rates"
#subtitle: "My subtitle if needed"
author: 
  - Maroosh Gillani
  - Aaron Liu
  - Jimmy Luc
thanks: "Code and data are available at: https://github.com/rex009x/us_birth_rates/tree/master/data . A replication of various aspects in this paper are available at: LINK ."
date: today
date-format: long
abstract: "Over the last couple of years, the American population has been witnessing a steady decline in their birth rates, as suggested by [@citeOriginalReplication]'s data from the American Economic Association. Through the analysis of multiple different factors, such as age and education, we notice a continual decline in birth rate starting from the year 2007. We explore how different social, cultural, and economic effects play a part in this worrying trend. This report sheds light on how the societal trends in the recent years have affected the American citizens in the last few decades, and how these patterns have played a part in this downward trend as people adapt to them."
format:
  pdf:
    fig-pos: "H"
toc: true
number-sections: true
bibliography: references.bib
---
```{r, include=FALSE, warning=FALSE, message=FALSE, echo=FALSE}
# workplace setup

@@

# acquire fig 5 data from nchs_cohort_analysis.dta
nchs_cohort_data <- 
  read_dta(here("data/nchs_cohort_analysis.dta"))
```

@@

```{r graph1, fig.cap="Trend in US Birth Rates from 1980 - 2020", echo=FALSE}
#| echo: false
#| message: false
#| warning: false

#| label: fig-1

@@

```{r graph2a, fig.cap="Trends in Birth Rates by Five-year Age Group", echo=FALSE}
#| echo: false
#| message: false
#| warning: false
#| label: fig-2

@@

```{r graph2b, fig.cap="Trends in Birth Rates by Mother's Level of Education (ages 20-44)", echo=FALSE}
# replicate figure 2d from the original paper
fig_2d_data_long %>% ggplot(aes(x = year, y = Birth_Rate, color = Education_Level)) +
  geom_line(size = 1) +
  labs(x = "Year", y = "Births per 1,000 women in \n relevant population subgroup", color = "Education Level") +
  scale_y_continuous(limits = c(0, 140), breaks = seq(0, 150, by = 20)) +
  scale_x_continuous(limits = c(1990, 2020), breaks = seq(1990, 2020, by = 5)) +
  geom_text(aes(x = 1993, y = 112), label = "No high school degree", color = "black", hjust = 0) +
  geom_text(aes(x = 1993, y = 82), label = "College graduate", color = "#f8766d", hjust = 0) +
  theme_classic() +
  geom_text(aes(x = 1993, y = 62), label = "High school graduate", color = "#00bfc4", hjust = 0) +
  geom_text(aes(x = 1993, y = 45), label = "Some college", color = "black", hjust = 0) +
  theme_classic() +
  theme(legend.position = "none")
```
```{r graph2c, fig.cap="Trends in Birth Rates by Parity (ages 15-44)", echo=FALSE}

@@

```{r graph3, fig.cap="Children Ever Born by Mother's Age and Birth Cohort", echo=FALSE}
#| echo: false
#| message: false
#| warning: false
#| label: fig-3

```
# References

>>>>

---
title: "A Deeper Analysis into the Trends and Factors of the US's Declining Birth Rates"
#subtitle: "My subtitle if needed"
author: 
  - Maroosh Gillani
  - Aaron Liu
  - Jimmy Luc
thanks: "Code and data are available at: https://github.com/rex009x/us_birth_rates/tree/master/data . A replication of various aspects in this paper are available at: LINK ."
date: today
date-format: long
abstract: "Over the last couple of years, the American population has been witnessing a steady decline in their birth rates, as suggested by [@citeOriginalReplication]'s data from the American Economic Association. Through the analysis of multiple different factors, such as age and education, we notice a continual decline in birth rate starting from the year 2007. We explore how different social, cultural, and economic effects play a part in this worrying trend. This report sheds light on how the societal trends in the recent years have affected the American citizens in the last few decades, and how these patterns have played a part in this downward trend as people adapt to them."
format:
  pdf:
    fig-pos: "H"
toc: true
number-sections: true
bibliography: references.bib
---
```{r, include=FALSE, warning=FALSE, message=FALSE}
# workplace setup

@@

# acquire fig 5 data from nchs_cohort_analysis.dta
# acquire figure 5 data from CDC SEER's agecomp-seer.dta
seer_cohort_data <- 
  read_dta(here("data/agecomp-seer.dta"))

nchs_cohort_data <- 
  read_dta(here("data/nchs_cohort_analysis.dta"))

# code to replicate figure 5 from: https://github.com/KCtt457/us-birth-rates-decline
# Process population data to aggregate by cohort
fig5_pop_data1 <- seer_cohort_data %>%
  select(1:32) %>%
  select(-c(stname)) %>%
  group_by(year) %>%
  summarise(across(everything(),sum)) # https://stackoverflow.com/questions/1660124/how-to-sum-a-variable-by-group

fig5_pop_data1 <- fig5_pop_data1 %>%
  pivot_longer(-c(year), names_to = "mage", values_to = "pop")

fig5_pop_data1 <- fig5_pop_data1 %>%
  mutate(mage = as.numeric(str_replace(mage, "fem", ""))) %>%
  mutate(cohort = year - mage) %>%
  mutate(cohort2 = case_when(cohort >= 1968 & cohort <=1972 ~ 1,
                             cohort >= 1973 & cohort <=1977 ~ 2,
                             cohort >= 1978 & cohort <=1982 ~ 3,
                             cohort >= 1983 & cohort <=1987 ~ 4,
                             cohort >= 1988 & cohort <=1992 ~ 5,
                             cohort >= 1993 & cohort <=1997 ~ 6)
         )

fig5_pop_data1 <- na.omit(fig5_pop_data1)
fig5_pop_data1 <- fig5_pop_data1 %>%
  group_by(cohort2, mage) %>%
  summarise(pop = sum(pop))

# Join dataframes for population and birth data
nchs_cohort_data <- full_join(nchs_cohort_data, fig5_pop_data1, by=c("mage", "cohort2")) %>%
  mutate(cohort_years = case_when(cohort2 == 1 ~ "1968-1972",
                                  cohort2 == 2 ~ "1973-1977",
                                  cohort2 == 3 ~ "1978-1982",
                                  cohort2 == 4 ~ "1983-1987",
                                  cohort2 == 5 ~ "1988-1992",
                                  cohort2 == 6 ~ "1993-1997")
         )

# Calculate birth rates
nchs_cohort_data <- nchs_cohort_data %>%
  mutate(brate=numbirth/pop*1000,
         age_20_24_year = case_when(cohort2 == 1 ~ 1992,
                                    cohort2 == 2 ~ 1997,
                                    cohort2 == 3 ~ 2002,
                                    cohort2 == 4 ~ 2007,
                                    cohort2 == 5 ~ 2012,
                                    cohort2 == 6 ~ 2017)
         )

# Calculate cumulative birth rates for each cohort
nchs_cohort_data <- nchs_cohort_data %>%
  group_by(cohort2) %>%
  mutate(cum_brate = cumsum(brate)/1000, 
         cohort_years = case_when(cohort2 == 1 ~ "1968-1972",
                                  cohort2 == 2 ~ "1973-1977",
                                  cohort2 == 3 ~ "1978-1982",
                                  cohort2 == 4 ~ "1983-1987",
                                  cohort2 == 5 ~ "1988-1992",
                                  cohort2 == 6 ~ "1993-1997")
         )

```

@@

```{r graph1, fig.cap="Trend in US Birth Rates from 1980 - 2020"}
#| echo: false
#| message: false
#| warning: false
#| label: fig-1

@@

```{r graph2a, fig.cap="Trends in Birth Rates by Five-year Age Group"}
#| echo: false
#| message: false
#| warning: false
#| label: fig-2a

@@

```{r graph2b, fig.cap="Trends in Birth Rates by Mother's Level of Education (ages 20-44)"}
#| echo: false
#| message: false
#| warning: false
#| label: fig-2b
# replicate figure 2d from the original paper
fig_2d_data_long %>% ggplot(aes(x = year, y = Birth_Rate, color = Education_Level)) +
  geom_line(size = 1) +
  labs(x = "Year", y = "Births per 1,000 women in \n relevant population subgroup", color = "Education Level") +
  scale_y_continuous(limits = c(0, 140), breaks = seq(0, 150, by = 20)) +
  scale_x_continuous(limits = c(1990, 2020), breaks = seq(1990, 2020, by = 5)) +
  geom_text(aes(x = 1993, y = 112), label = "No high school degree", color = "black", hjust = 0) +
  geom_text(aes(x = 1993, y = 82), label = "College graduate", color = "#f8766d", hjust = 0) +
  theme_classic() +
  geom_text(aes(x = 1993, y = 62), label = "High school graduate", color = "#00bfc4", hjust = 0) +
  geom_text(aes(x = 1993, y = 45), label = "Some college", color = "black", hjust = 0) +
  theme_classic() +
  theme(legend.position = "none")
```

```{r graph2c, fig.cap="Trends in Birth Rates by Parity (ages 15-44)"}
#| echo: false
#| message: false
#| warning: false
#| label: fig-2c

@@

```{r graph3, fig.cap="Children Ever Born by Mother's Age and Birth Cohort"}
#| echo: false
#| message: false
#| warning: false
#| label: fig-3

# replicate figure 5 from the original paper
# code to replicate figure 5 from: https://github.com/KCtt457/us-birth-rates-decline
nchs_cohort_data %>%
  ggplot(aes(x = mage,
             y = cum_brate,
             color = cohort_years)) +
  geom_line(size = 1) +
  labs(x = "Mother's age", y = "Children ever born", color = "Birth cohorts") +
  scale_y_continuous(limits=c(0, 2.5), breaks=seq(0, 2.5, by = .5)) +
  scale_x_continuous(limits=c(15, 44), breaks=seq(15, 44, by = 1)) +
  theme_classic() + 
  theme(legend.background = element_blank(), legend.box.background = element_rect(), legend.position = c(0.85,0.4), axis.text.x = element_text(angle = 45, hjust = 1))
```
# References
